Efficient simulation of individual-based population models

### 2024-03-25 article.qmd
update

In @sec-ibm, we introduce the mathematical framework that characterizes the class of Stochastic Individual-Based Models (IBMs) that can be implemented in the `IBMPopSim` package. In particular, a general pathwise representation of IBMs is presented. The population dynamics is obtained as the solution of an SDE driven by Poisson measures, for which we obtain existence and uniqueness results in @thm-eq-Z. Additionally, a succinct overview of the package is provided. 
In @sec-simulation the two main algorithms for simulating the population evolution of an IBM across the interval $[0, T]$ are detailed. 
In @sec-package we present the main functions of the `IBMPopSim` package, which allow for the definition of events and their intensities, the creation of a model, and the simulation of scenarios. 
Two examples are detailed in @sec-insurance-portfolio and sec-example-interaction, featuring applications involving an heterogeneous insurance portfolio characterized by entry and exit events, and an age and size-structured population with intricate interactions.

@@

2. In the presence of competition between individuals, the death intensity of an individual $I$ also depend on other individuals $J$ in the population. For example, if $I=(\tau^b,\tau^d, x)$, with $x$ its size, then we can have: 
$$
    W^d(t,I,J) = (x_J - x)^+ \mathbf{1}_{\{\tau^d_J > t\}}, \quad \forall \; J=(\tau^b_J,\tau^d_J , x_J).
$$ {#eq-ex-interation}
This can be interpreted as follows: if the individual $I$ meets randomly an individual $J$ alive at time $t$, and of bigger size $x_J > x$, then he can die at the intensity $x_J-x$. If $J$ is smaller than $I$, then he cannot kill $I$. The bigger is the size $x$ of $I$, the lower is his death intensity $\lambda^d_t(I,Z_t)$ defined by 
$$
    \lambda^d_t(I,Z_t) = \sum_{\genfrac{}{}{0pt}{2}{J\in Z_t,}{x_J > x}} (x_J -x)\mathbf{1}_{\{\tau^d_J > t\}}.
$$

@@

The proof of @thm-interaction is detailed in the Appendix @prf-no-interaction.

@@

@@

An event $e\in \mathcal{E}$ (see @eq-no-interaction) has an intensity of the form $\lambda^e(t, I)$ which depends only on the individual `I` and time. Events with such intensity are created using the function
```{R}
#| eval: false
mk_event_individual(type = "TYPE", 
                    name = "NAME",
                    intensity_code = "INTENSITY", ...)
```
The `intensity_code` argument is a character string containing few lines of `C++` code describing the intensity function $\lambda^e(t, I)$. The intensity value has to be stored in a variable called `result` and the available variables for the intensity code are given in @tbl-intensity-variables.

@@

The argument `interaction_type`, set by default at `random`, is the algorithm choice for simulating the model. When `interaction_type=full`, the simulation follows @cnj-no-interaction, `interaction_type=random` it follows @cnj-interaction.
In most cases, the `random` algorithm is much faster than the `full` algorithm, as we illustrate for instance in @sec-example-interaction , where we observe the gain of a factor of 40 between the two algorithms, on a set of standard parameters. This allows in particular to explore parameter sets that give larger population sizes, without reaching computation times that explode.

@@

# Appendix {.appendix .unnumbered}

@@

>>>>

In @sec-ibm, we introduce the mathematical framework that characterizes the class of Stochastic Individual-Based Models (IBMs) that can be implemented in the `IBMPopSim` package. In particular, a general pathwise representation of IBMs is presented. The population dynamics is obtained as the solution of an SDE driven by Poisson measures, for which we obtain existence and uniqueness results in @thm-eq-Z. Additionally, a succinct overview of the package is provided. 
In @sec-simulation the two main algorithms for simulating the population evolution of an IBM across the interval $[0, T]$ are detailed. 
In @sec-package we present the main functions of the `IBMPopSim` package, which allow for the definition of events and their intensities, the creation of a model, and the simulation of scenarios. 
Two examples are detailed in @sec-insurance-portfolio and @sec-example-interaction, featuring applications involving an heterogeneous insurance portfolio characterized by entry and exit events, and an age and size-structured population with intricate interactions.

@@

2. In the presence of competition between individuals, the death intensity of an individual $I$ also depend on other individuals $J$ in the population. For example, if $I=(\tau^b,\tau^d, x)$, with $x$ its size, then we can have: 
$$
    W^d(t,I,J) = (x_J - x)^+ \mathbf{1}_{\{\tau^d_J > t\}}, \quad \forall \; J=(\tau^b_J,\tau^d_J , x_J).
$$ {#eq-ex-interaction}
This can be interpreted as follows: if the individual $I$ meets randomly an individual $J$ alive at time $t$, and of bigger size $x_J > x$, then he can die at the intensity $x_J-x$. If $J$ is smaller than $I$, then he cannot kill $I$. The bigger is the size $x$ of $I$, the lower is his death intensity $\lambda^d_t(I,Z_t)$ defined by 
$$
    \lambda^d_t(I,Z_t) = \sum_{\genfrac{}{}{0pt}{2}{J\in Z_t,}{x_J > x}} (x_J -x)\mathbf{1}_{\{\tau^d_J > t\}}.
$$

@@

The proof of @thm-interaction is detailed in the Appendix @sec-prf-thm-interaction.

@@

## Population {#sec-package-population}

@@

An event $e\in \mathcal{E}$ (see @eq-intensity-no-interaction) has an intensity of the form $\lambda^e(t, I)$ which depends only on the individual `I` and time. Events with such intensity are created using the function
```{R}
#| eval: false
mk_event_individual(type = "TYPE", 
                    name = "NAME",
                    intensity_code = "INTENSITY", ...)
```
The `intensity_code` argument is a character string containing few lines of `C++` code describing the intensity function $\lambda^e(t, I)$. The intensity value has to be stored in a variable called `result` and the available variables for the intensity code are given in @tbl-intensity-variables.

@@

The argument `interaction_type`, set by default at `random`, is the algorithm choice for simulating the model. When `interaction_type=full`, the simulation follows @alg-no-interaction, `interaction_type=random` it follows @alg-interaction.
In most cases, the `random` algorithm is much faster than the `full` algorithm, as we illustrate for instance in @sec-example-interaction , where we observe the gain of a factor of 40 between the two algorithms, on a set of standard parameters. This allows in particular to explore parameter sets that give larger population sizes, without reaching computation times that explode.

@@

# Appendix {.appendix}

@@

### Proof of @lem-non-explosion {#sec-prf-lem-non-explosion}

The proof is obtained using pathwise comparison result, generalizing those obtained in [@KaaElK20].

::: {.proof}
TODO
:::

### Alternative pathwise representation 
TODO

## Proof of @thm-interaction {#sec-prf-thm-interaction}
